### 2004-07-13 Documentation/Paper/paper.tex
*** empty log message ***

\newcommand{\JHFSK}{{\sc JHF-SK}}
\newcommand{\NUMI }{{\sc NuMI}}
\newcommand{\ReactorI}{{\sc Reactor-I}}
\newcommand{\ReactorII}{{\sc Reactor-II}}
\newcommand{\JHFHK}{{\sc JHF-HK}}
\newcommand{\NuFactI}{{\sc NuFact-I}}
\newcommand{\NuFactII}{{\sc NuFact-II}}

@@

\begin{abstract}
This is the abstract.
\end{abstract}

@@

@@

\bi
 \item 
  \figu{GLOBES} and main features
 \item
  Physics basics: For what purpose? What are difficulties
 \item
  Detector simulation: What happens there?
 \item
  References for work done with \GLOBES\
 \item
  Parts of \GLOBES\
 \item 
  \figu{GLOBES} and main features:
  \bi
   \item
    3-flavor analysis
   \item
    Matter density profile+uncertainty
   \item
    Full detector simulation
   \item
    Up to 32 experiments simultaneously
   \item
    Full n-parameter correlation
   \item
    Localization of degeneracies
  \ei
\ei

@@

\begin{figure}[t]
\begin{center}
\begin{tabular}{p{5cm}p{5cm}p{5cm}}
\includegraphics[width=5cm]{AEDL1} &
\includegraphics[width=5cm]{SignalBackground} &
\includegraphics[width=5cm]{Rules} \\
\end{tabular}
\end{center}
\caption{\label{fig:aedl} The most important components of \AEDL :
Channels, rules, and experiments.}
\end{figure}

@@

\bi
\item
 Complexity of projection: $\mathcal{O}(n^7)$ (incl. matter density) API
 $\rightarrow$ Polynomical, but large effort;
 Approach with Local minimizer \ra\ $\mathcal{O}(1000)$ steps
\item
 Running time of detector simulation: (each step above)
 bins times sampling points + sampling points times matter profile steps.
 Normally dominated by first summand; for many matter profile steps
 second dominates
\item
 Computational effort: Estimates (time)
\item
 Pre-computation of smearing matrices, Spezielles
 Format bei dem Nullen nicht mitgerechnet;
  oscillation probabilities nur einmal gleichzeitig fuer alls
  Oszillationskanaele und alle ENergien berechnet
\ei

\section{Use of program}

\bi
\item
 Whole package
\item
 Shared library, globes command, Examples, Experiment protoypes
\item
 Manual \cite{Manual}
\ei

@@

>>>>

\newcommand{\MINOS}{{\sf MINOS}}
\newcommand{\ICARUS}{{\sf ICARUS}}
\newcommand{\OPERA}{{\sf OPERA}}
\newcommand{\JHFSK}{{\sf JHF-SK}}
\newcommand{\NUMI }{{\sf NuMI}}
\newcommand{\ReactorI}{{\sf Reactor-I}}
\newcommand{\ReactorII}{{\sf Reactor-II}}
\newcommand{\JHFHK}{{\sf JHF-HK}}
\newcommand{\NuFactI}{{\sf NuFact-I}}
\newcommand{\NuFactII}{{\sf NuFact-II}}

@@

\begin{abstract}
We present the \GLOBES\ (``General Long Baseline Experiment Simulator'')
software package, which allows the simulation of future long-baseline
neutrino oscillation experiments. One part of the software is the
abstract experiment definition language to define experiments with
beam and full detector simulations as accurate as possible. For the
definition of experiments, a new transparent building block concept is introduced. In addition, a binary provides the possibility to test
new experiment definitions quickly.
%
Another
part of \GLOBES\ is the user's interface, which provides probability, rate, and $\Delta \chi^2$ information for a given experiment or any
combination of up to 32 experiments. Especially, the
$\Delta \chi^2$ functions allow a simulation with statistics only,
systematics, correlations, and degeneracies. In particular, \GLOBES\
can handle the full multi-parameter correlation among the oscillation
parameters, external input, and matter density uncertainties. Eventually,
many systematics options are provided, such as normalization and energy
calibration errors, or the choice between spectral or total rate 
information. 
\end{abstract}

@@

Neutrino oscillations are now established as the leading flavor
transition mechanism for neutrinos. By the existing experiments, 
two oscillation
frequencies have been found: The fast ``atmospheric'' and the
slow ``solar'' oscillations, which are driven by the respective
mass squared differences.  In addition, there could be interference
effects between these two oscillations, provided that the coupling given by the small mixing angle $\theta_{13}$ is large enough. Such interference
effects include, for example, leptonic CP violation.  In order to
test the unknown oscillation parameters, \ie, the mixing angle $\theta_{13}$, the leptonic CP phase, and the
neutrino mass hierarchy, future long-baseline experiments are proposed.
These experiments send an artificial neutrino beam to a detector, or
detect the neutrinos produced by a nuclear fission reactor. However,
instrinsic degenerate solutions in the neutrino oscillation
probabilities affect these measurements and require optimization
strategies for the experiments to maximally exploit complementary
information. Therefore, a modern complete experiment simulation
does not only need to have a beam and detector simulations, which are
as accurate as possible, but also powerful tools to analyze correlations
and degeneracies. In this study, we present the \GLOBES\ software package
as such a tool.

\GLOBES\ (``General Long Baseline Experiment Simulator'') is a flexible
software tool to simulate neutrino oscillation long-baseline experiments using a complete three-flavor analysis. On the
one hand, it contains a comprehensive abstract experiment definition
language, which allows to describe most classes of long baseline experiments
at an abstract level. On the other hand, it provides a C-library to 
process the experiment information in order to obtain oscillation
probabilities, rate vectors, and $\Delta \chi^2$-values (\cf, \figu{GLOBES}). In addition, it provides a binary to test experiment
definitions very quickly, before they are used by the application software.
Currently, \GLOBES\ is available for Linux. Since the source code is included, the modifications to other operating systems should be doable.

\GLOBES\ allows to simulate experiments with stationary neutrino point sources, where each experiment is assumed to have only one neutrino source.
Such experiments are neutrino beam experiments and reactor experiments. 
Geometrical effects of a source distribution, such as in the sun or the 
atmosphere, can not be described. In addition, sources with a physically 
significant time dependencies  can not be studied, such as  supernov\ae. It 
is, however, possible to simulate beams with bunch structure, since the 
time dependence of the neutrino source is physically only important to suppress backgrounds.

@@

On the experiment definition side, either built-in neutrino fluxes
(\eg, neutrino factory) or arbitrary fluxes can be used. Similarly,
arbitrary cross sections, energy dependent efficiencies, the
energy resolution function, the considered oscillation channels, 
backgrounds, and many other features can be specified. 
For the systematics, energy
normalization and calibration errors can be simulated. Note that
the energy ranges and windows, as well as the bin widths can be
(almost) arbitrarily chosen, which means that variable bin
widths are allowed. Together with \GLOBES\ comes a number of
pre-defined experiments in order to demonstrate the capabilities
of \GLOBES\ and to provide prototypes for new experiments.
Examples for these prototypes are the \MINOS , \ICARUS , and
\OPERA\ simulations from \Ref~\cite{Huber:2004ug}, the
\JHFSK\ and \NUMI\ superbeam simulations from \Refs~\cite{Huber:2002mx,
Huber:2002rs}, the \JHFHK\ superbeam upgrade simulation from
\Ref~\cite{Huber:2002mx}, the neutrino factory simulations from
\Ref~\cite{Huber:2002mx}, and the reactor experiment simulations from
\Ref~\cite{Huber:2003pm}. Other projects using the \GLOBES\ 
software include \Refs~\cite{Huber:2003ak,Ohlsson:2003ip,Winter:2003ye,Antusch:2004yx}.

With the C-library, one can extract the $\Delta \chi^2$ for all defined 
oscillation channels for an experiment or any combination of experiments.
Of course, also low-level information, such as oscillation
probabilities or event rates, can be obtained. \GLOBES\ includes the
simulation of neutrino oscillations in matter with arbitrary matter 
density profiles, as well as it allows to simulate the matter density
uncertainty. As one of the most
advanced features of \GLOBES , it provides the technology to 
project the $\Delta \chi^2$, which is a function of all oscillation
parameters, onto any subspace of parameters by local minimization. 
This approach allows the inclusion of multi-parameter-correlations,
where external input (\eg, from solar parameters) can be imposed, too.
Applications of the projection mechanism include the projections onto the $\stheta$-axis and the $\stheta$-$\deltacp$-plane. In addition, all oscillation parameters can be kept free to precisely localize 
degenerate solutions.

@@

\begin{figure}[t]
\begin{center}
\begin{tabular}{p{6.3cm}p{4.3cm}p{4.3cm}}
\includegraphics[width=6.5cm]{AEDL1} &
\includegraphics[width=4.5cm]{SignalBackground} &
\includegraphics[width=4.5cm]{Rules} \\
\end{tabular}
\end{center}
\caption{\label{fig:aedl} The most important components of \AEDL :
Channels, rules, and experiments.}
\end{figure}

@@

The \GLOBES\ software only uses polynomial algorithms, which are,
however, suffering from the large number of dimensions.
The $\Delta \chi^2$ computation basically consists of two steps:
The systematics $(\Delta \chi^2)_{\mathrm{sys}}$ computation with {\tt glbChiSys},
and the projection algorithm which uses {\tt glbChiSys} many times. 
The overall computation time is obtained as the product of these two steps.

The running time of the systematics $(\Delta \chi^2)_{\mathrm{sys}}$ 
is 
\begin{equation}
\mathrm{t_{sys}} = \mathcal{O} \left( N_{\mathrm{Sampling}} \times N_{\mathrm{Bins}}  \, + \, N_{\mathrm{Sampling}} \times N_{\mathrm{Layers}} \right) ,
\end{equation}
since the oscillation probabilities for all channels and energies
are only computed once and stored in a list. Here $N_{\mathrm{Sampling}}$ is the number of sampling points for the evaluation of \equ{events_bin}, $N_{\mathrm{Bins}}$ is the number of energy bins,
and $N_{\mathrm{Layers}}$ is the number of matter density layers.\footnote{Note that we are here using the \AEDL\ format for these
parameters.}. Thus, whenever the
number of matter density steps is small, the product of sampling points
and bins dominates. In practice, the $N_{\mathrm{Sampling}} \times N_{\mathrm{Bins}}$ energy smearing matrix is already computed at 
the experiment initialization in order to save running time. In addition,
since it contains a lot of zeros, \GLOBES\ uses a special format to
avoid looping over the zeros.

The running time for the projected $\Delta \chi^2$ is, for a grid based method,
\begin{equation}
 \mathrm{t_{proj}^{Grid}} = \mathcal{O}\left( n^{6+N_{\mathrm{exp}}}\right) \, \times \, \mathrm{t_{sys}},
\end{equation}
where $n$ is the number of steps to be evaluated for each oscillation
parameter, and $N_{\mathrm{exp}}$ is the number of experiments
(since the matter densities of all experiments appear as parameters). 
Though it is, in principle, possible to use such a method with \GLOBES ,
it is impractiable in most cases, since it would involve at least 
billions of
steps for the complete multi-parameter correlation. The local $6+N_{\mathrm{exp}}$-dimensional minimization reduces this effort
to 
\begin{equation}
\mathrm{t_{proj}^{Minimizer}} = \mathcal{O}(1000) \, \times \,
N_{\mathrm{Deg}} \,  \times \, \mathrm{t_{sys}}
\end{equation}
steps, where $N_{\mathrm{Deg}}$ is the number of (disconnected) 
local minima at the chosen confidence level, \ie, the number of
degeneracies.

Eventually, \GLOBES\ needs about $10$ to $15$ seconds to compute
a projected $\Delta \chi^2$ on a modern Pentium machine. Therefore,
more sophisticated applications, such as a two dimensional plot
for CP violation as function of the simulated values of $\stheta$ and
$\deltacp$ can be obtained in several hours.

\section{Use of package}

The \GLOBES\ package~\cite{Globes} is a tar-Ball which consists of the
C-library for the user's interface, a binary to test
\AEDL\ files, example files, 
experiment prototypes with supporting flux and cross 
section files, and other 
supporting files for installation and compilation.
Currently, the C-library is only provided for Linux.
In addition, the user's interface and \AEDL\ manual~\cite{Manual} 
can be obtained separately. Upon extracting the tar-ball,
the \GLOBES\ user's interface library needs to be configured, 
compiled, and installed as 
described in the {\tt INSTALL} file. In principle, it can be
compiled as a static library or a dynamically-linked shared library
(which can be installed with superuser rights or only used in one's
home directory). Then it can be linked against any C program
containing the application software, such as the examples
from the manual. \AEDL\ files, such as the experiment protoypes,
can be edited with any text editor, in order to be loaded by the
users's interface later. In addition, a binary is provided to
test and create experiment descriptions. For example, event rate
information can be quickly provided to adjust the neutrino flux
normalization. For further information, we refer
at this place to the \GLOBES\ manual~\cite{Manual}.

@@

In summary, the \GLOBES\ software package provides powerful tools
to do a full three-flavor analysis of future long-baseline experiments
including systematics, correlations, and degeneracies. In addition,
the abstract experiment definition language allows to define
experiments at an abstract level as accurate as possible. 
Some of the major strengths of \GLOBES\ are the ability to quickly
define new experiments, the potential to take into account the
full multi-parameter correlation, and the possibilities to include
external input and matter density uncertainties.

We conclude that the \GLOBES\ software has two major target groups:
Experimentalists, who want to quickly evaluate the physics potential
of their setups, and theorists, who want to test new ideas or
strategies with pre-defined experiments. Especially, the separation
between experiment descriptions as simple text files (together
with supporting files) and the application software should allow an
efficient and lively interaction between those two target groups.

\subsection*{Acknowledgments}

We would like to thank Martin Freund and Andrea Romanino ???
for initiating this project many years ago with one of the first
neutrino factory simulations. In addition, we would like to thank
Marc Rolinec for the help to translate the experiment descriptions
into \AEDL , and for creating the illustrations in this manuscript.
Finally, thanks to all the people who have been pushing this project
for many years, to the ones who have been continuing asking for the publication of the software, and the referees of several of our
papers for suggestions which lead to improvements in the software.
